Who Cares About Crackdowns? Exploring the Role of Trust in Individual Philanthropy

### 2020-09-29 manuscript/manuscript.Rmd
Updates with new simulation results

---
title: "Who Cares About Crackdowns? Exploring the Role of Trust in Individual Philanthropy"
short-title: "Who Cares About Crackdowns?"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Marc Dotson
  affiliation: Brigham Young University
  url: https://marriottschool.byu.edu/directory/details?id=50683
  email: marc_dotson@byu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
date: "September 24, 2020"
published: Working paper.
git-repo: https://example.com
abstract: >-
  Over the past two decades, international non-governmental organizations (INGOs) have faced increasing legal scrutiny and repression abroad. As a result, official aid to INGOs in repressive countries has reduced. NGOs thus need access to a diverse array of funds and may need to increase fundraising from private donors. Which demographic and experiential factors influence the decision of individuals to donate to legally besieged NGOs? We investigate this question using a market simulation based on a nationally representative conjoint experiment conducted on individuals with a history of charitable giving. We find that trust in political institutions and charities is enormously influential in shaping donor preferences. Donor profiles with high levels of social trust are likely to donate to human rights INGOs with friendly relationships with host governments, but that support evaporates if INGOs face criticism or crackdown. In contrast, donor profiles with lower levels of social trust prefer to donate to organizations that do face criticism or crackdown abroad. Our findings have practical implications for INGOs that aim to raise funds from the general public amid closing civic space. 
reference-section-title: References
toc: false
mainfont: Lora
sansfont: IBM Plex Sans
---

@@

Existing scholarship on the determinants of individual-level philanthropy to domestic causes is robust across a variety of contexts. But the dynamics of *international* giving to NGOs are under-examined in the literature on individual donor attributes and preferences.[^exceptions] Motivating individuals to donate to international causes is hard as the number of recipients are large and far removed from the donor [@CasaleBaumann:2015, 100]. And research on the determinants of philanthropy for international causes—particularly towards organizations working in challenging environments—is still in its infancy due to the absence of expansive and high-quality data [@greenhill2013]. While previous work has looked at how organizational attributes of nonprofits affect donors' willingness to donate internationally [@todo:CITEOURPAPERTODIFFERENTIATE), we have yet to uncover demographic traits and experiences that motivate or deter individuals to donate during shrinking civic space.

@@

We begin by looking at differences in donation preferences across a range of basic demographic characteristics. The three panels of @fig:demographics shows the average predicted market share for different organizational issue areas, relationships with governments, and funding sources. Though we varied gender across the different personas, gender has no significant effect on any of the simulated outcomes—men and women follow nearly identical patterns across all organizational and individual characteristics. As such, in @fig:demographics, we only vary income. Personas with high income earn \$100,000 a year, while those with less income earn \$50,000 annually.

![Average predicted donation market shares across all personas, segmented by individual persona income and organizational issue area, relationship with host government, and funding sources](output/figures/income-all.pdf){#fig:demographics}

Wealthier personas respond better to human rights organizations over refugee assistance and environmental advocacy organizations, while personas that earn less money tend to prefer donating to refugee and environmental organizations. As we explore below, this is likely because income correlates with interest in public affairs and political activity—individuals who are wealthier arguably have more time and resources to follow the news and be continuously engaged in their communities and in other activist causes.

There are also sizable income-based differences across different NGO host-country relationships. Wealthier donors are more likely to prefer donating to organizations that are criticized by their host governments. This falls in line with their preference for human rights organizations, which have more contentious missions and are more likely to receive government criticism. This support, however, appears to evaporate as civil society crackdown intensifies. High income personas are the least likely demographic to prefer donating to NGOs under crackdown. Their lower income counterparts, on the other hand, are generally unfazed by NGOs' relationships with government, maintaining the same average market share across all three types of host country relationships. Finally, as seen in the right panel of @fig:demographics, there are no sizable differences in donation preferences by income across different forms of NGO funding. Income thus appears to influence preferences particularly for NGOs facing criticism and crackdown abroad—wealthier donors are perhaps fair-weather friends that are more likely to pull their support when an organization's work becomes more difficult.

*(TODO: I'll need ≈150 words to describe education and religiosity results)*

@@

@fig:issue-social, @fig:relationship-social, and @fig:funding-social show the results of the market simulation across these three dimensions of persona characteristics, varying public affairs knowledge, political ideology, and social trust. Personas respond differently to organizational issue areas. In @fig:issue-social, those who follow the news and travel abroad are more likely to prefer donating to human rights organizations than their less knowledgeable and less traveled counterparts. Personas who do not follow the news or travel show a greater preference for emergency response and refugee relief organization, similar to the income effects seen previously. There are no strong ideological effects across issue areas. Both conservative and liberal personas tend to follow similar patterns of preference across NGO issue.

@@

@fig:relationship-social shows how personas respond to different NGO–host government relationships. Public affairs knowledge and experience has little effect on donor preferences, as the average market shares are roughly identical across the two columns of the figure. Persona preferences toward friendly, criticized, and legally suppressed organizations do not change much across different ideological views. The strongest, most overriding factor in determining donor preferences when considering an organization's relationship with its host country is social trust. Personas with high levels of social trust are overwhelmingly most likely to donate to organizations with friendly relationships with their host governments, while those with low levels of trust avoid friendly NGOs almost entirely.

This stark difference in attitudes towards friendly NGOs likely reflects the underlying levels of trust in the model. For those who trust charities to do the right thing and have a positive influence on society, a friendly relationship with a host government is a manifestation of that belief and a sign that the organization is living up to its ideal. Once the organization faces criticism or crackdown, however, support deteriorates rapidly, since such a crackdown could be perceived as a violation of the persona's trust in the charity—that is, the organization must be doing something wrong to deserve criticism. Personas with low levels of trust, on the other hand, eschew friendly organizations entirely and are far more willing to donate to organizations facing criticism and crackdown, likely because they do not assume that charities must always be above reproach. The global crackdown on NGOs could thus counterintuitively sour NGOs' most trusting donors.

@@

More importantly, we find that trust in political institutions and in charitable organizations matters substantially in shaping donor preferences. Personas with high levels of social and charitable trust prefer donating to human rights organizations that have friendly relationships with their host governments, while personas with low trust are drawn more to emergency response and refugee relief organizations with unfriendly relationships with their host governments. Personas with low social trust rarely prefer to donate to any type of organization with friendly host-country relationships, while personas with high social trust tend to avoid organizations with unfriendly relationships. This could imply that donors that are the most trusting and involved might be fair-weather friends that drop their support when charities face government criticism and crackdown.

The above results have important implications for NGOs as they navigate a shrinking civic space and consider the use of different frames to tailor their fundraising appeals to individual donors. Our survey finds that at least for some kinds of individuals, NGOs may benefit from publicizing when they are targets of government crackdown, including that information in their fundraising appeals. However, the bigger challenge may be addressing concerns of donors with low levels of trust. More broadly, studies from a range of countries show that levels of giving are lowest for refugee and asylum organizations, as well as for overseas and development organizations. Rather, donors giving to international causes tend to support international disaster relief and religious groups [@RobsonHart:2020]. However, the former are precisely the organizations facing crackdown. More research is needed to answer how these organizations can appeal to broader array of potential foreign donors to increase levels of charitable-giving towards them.

@@

[^model-details]:	Specifically,

    $$
    \begin{aligned}
    \beta &\sim \operatorname{Multivariate} \mathcal{N}(Z \Gamma, \xi) \\
    y &\sim \operatorname{Multinomial\ logit}(X \beta, \varepsilon)
    \end{aligned}
    $$

    where $y$ = which alternative the respondent chooses to donate to, $X$ = design matrix of attribute levels (i.e. organizations, issue areas, organizational practices, funding sources, and government relations), $\beta$ = latent individual preferences for the attribute levels, $Z$ = matrix of individual-level covariates (i.e. demographics, political knowledge, attitudes towards charity, etc.), $\Gamma$ = matrix of coefficients mapping individual-level covariates onto the latent individual-level preferences, and $\varepsilon$ and $\xi$ = errors.

>>>>

---
title: "Who Cares About Crackdowns? Exploring the Role of Trust in Individual Philanthropy"
short-title: "Who Cares About Crackdowns?"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Marc Dotson
  affiliation: Brigham Young University
  url: https://marriottschool.byu.edu/directory/details?id=50683
  email: marc_dotson@byu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
date: "September 24, 2020"
published: Working paper.
git-repo: https://example.com
abstract: >-
  Over the past two decades, international non-governmental organizations (INGOs) have faced increasing legal scrutiny and repression abroad. As a result, official aid to INGOs in repressive countries has reduced. NGOs thus need access to a diverse array of funds and may need to increase fundraising from private donors. Which demographic and experiential factors influence the decision of individuals to donate to legally besieged NGOs? We investigate this question using a market simulation based on a nationally representative conjoint experiment conducted on individuals with a history of charitable giving. We find that trust in political institutions and charities is enormously influential in shaping donor preferences. Donor profiles with high levels of social trust are likely to donate to human rights INGOs with friendly relationships with host governments, and that support hold steady if INGOs face criticism or crackdown. In contrast, donor profiles with lower levels of social trust prefer to donate to organizations that do not face criticism or crackdown abroad. Our findings have practical implications for INGOs that aim to raise funds from the general public amid closing civic space. 
reference-section-title: References
toc: false
mainfont: Lora
sansfont: IBM Plex Sans
---

@@

Existing scholarship on the determinants of individual-level philanthropy to domestic causes is robust across a variety of contexts. But the dynamics of *international* giving to NGOs are under-examined in the literature on individual donor attributes and preferences.[^exceptions] Motivating individuals to donate to international causes is hard as the number of recipients are large and far removed from the donor [@CasaleBaumann:2015, 100]. And research on the determinants of philanthropy for international causes—particularly towards organizations working in challenging environments—is still in its infancy due to the absence of expansive and high-quality data [@greenhill2013]. While previous work has looked at how organizational attributes of nonprofits affect donors' willingness to donate internationally [@ChaudhryHeiss:2020], we have yet to uncover demographic traits and experiences that motivate or deter individuals to donate during shrinking civic space.

@@

We begin by looking at differences in donation preferences across a range of basic demographic characteristics. The four panels of @fig:income show the average predicted market share for different organizational issue areas, funding sources, and relationships with host governments across different levels of persona income, with two different cutpoints for high and low income.

![Average predicted donation market shares across all personas, segmented by individual persona income and organizational issue area, relationship with host government, and funding sources](output/figures/income-all.pdf){#fig:income}

In general, personas respond better to human rights and refugee assistance organizations and environmental advocacy organizations across both levels of income, though wealthier personas show a slight preference towards human rights and refugee organizations. As we explore below, this is likely because income correlates with interest in public affairs and political activity—individuals who are wealthier arguably have more time and resources to follow the news and be continuously engaged in their communities and in other activist causes. There are no sizable differences in donation preferences by income across different forms of NGO funding—both levels of income tend to prefer organizations funded by government grants over small donors.

An organization's relationship with its host government does influence market shares across levels of income, but the size of this difference depends on the magnitude of income. When splitting income based on the US median income (\$61,372), there is no difference between income levels. In general, personas prefer to donate to organizations with friendly relationships with their host governments and avoid those that are criticized or under crackdown. However, instead of looking at personas with below- and above-median incomes, we can simulate a subset of wealthier donors that earn at least \$100,000 annually. These personas are more likely to prefer donating to organizations that are criticized by their host governments. This falls in line with their preference for human rights organizations, which have more contentious missions and are more likely to receive government criticism. This support, however, appears to evaporate as civil society crackdown intensifies. High income personas are the least likely demographic to prefer donating to NGOs under crackdown. Their lower income counterparts, on the other hand, are generally unfazed by NGOs' relationships with government, maintaining the same average market share across all three types of host country relationships. Patterns in preferences for issue area and funding sources remain constant across both methods of categorizing income—crackdown only has an effect on very high income personas. Income thus appears to influence preferences particularly for NGOs facing criticism and crackdown abroad—wealthy donors are perhaps fair-weather friends that are more likely to pull their support when an organization's work becomes more difficult.

While education and religiosity are both important drivers of individual philanthropy, neither are associated with any sizable differences in persona donation patterns. In general, both high school and college graduates *and* personas who rarely attend and who often attend religious services follow the same patterns in preferences shown in @fig:income: personas prefer human rights and refugee relief organizations, organizations funded by government grants, and organizations with friendly relaitonships with their host governments (see @fig:education and @fig:religiosity in the appendix.)

@@

@fig:issue-social and @fig:relationship-social show the results of the market simulation across these three dimensions of persona characteristics, varying public affairs knowledge, political ideology, and social trust. Personas respond differently to organizational issue areas. In @fig:issue-social, those who follow the news and travel abroad are more likely to prefer donating to human rights organizations than their less knowledgeable and less traveled counterparts. Personas who do not follow the news or travel show a greater preference for emergency response and refugee relief organization, similar to the income effects seen previously. There are no strong ideological effects across issue areas. Both conservative and liberal personas tend to follow similar patterns of preference across NGO issue.

@@

@fig:relationship-social shows how personas respond to different NGO–host government relationships. Public affairs knowledge and experience has little effect on donor preferences, as the average market shares are roughly identical across the two columns of the figure. Persona preferences toward friendly, criticized, and legally suppressed organizations do not change much across different ideological views. The strongest, most overriding factor in determining donor preferences when considering an organization's relationship with its host country is social trust.

Personas across both levels of social trust are most likely to donate to donate to organizations with friendly relationships with their host governments. However, those with higher trust are less likely to be deterred by poor host-country relationships, while those with low trust tend to not donate to organizations facing criticism or crackdown abroad. A friendly relationship with a host government is a manifestation of a belief that charities behave properly and exert a positive influence on society. Once an organization faces criticism or crackdown, however, support deteriorates rapidly among personas with low social trust, since such a crackdown could be perceived as a violation of the persona's trust in the charity—that is, the organization must be doing something wrong to deserve criticism. Personas with high levels of trust, on the other hand, seem to stick with legally besieged organizations and show them the benefit of the doubt. The global crackdown on NGOs thus sours NGOs' least trusting donors.

@@

More importantly, we find that trust in political institutions and in charitable organizations matters substantially in shaping donor preferences. Personas with high levels of social and charitable trust prefer donating to human rights organizations that have friendly relationships with their host governments, while personas with low trust are drawn more to emergency response and refugee relief organizations with unfriendly relationships with their host governments. Personas with low social trust tend to be scared off by negative host-country relationships and strongly prefer to donate to organizations that are friendly with their host governments, while personas with high social trust are more willing to stick with legally besieged organizations. This could imply that donors that are the most trusting and involved could remain supportive when charities face government criticism and crackdown.

The above results have important implications for NGOs as they navigate a shrinking civic space and consider the use of different frames to tailor their fundraising appeals to individual donors. Our survey finds that at least for some kinds of individuals, NGOs may benefit from publicizing when they are targets of government crackdown, including that information in their fundraising appeals. **TODO THIS IS DIFFERENT NOW** However, the bigger challenge may be addressing concerns of donors with low levels of trust. More broadly, studies from a range of countries show that levels of giving are lowest for refugee and asylum organizations, as well as for overseas and development organizations. Rather, donors giving to international causes tend to support international disaster relief and religious groups [@RobsonHart:2020]. However, the former are precisely the organizations facing crackdown. More research is needed to answer how these organizations can appeal to broader array of potential foreign donors to increase levels of charitable-giving towards them.

@@

[^model-details]: See the appendix for technical details about the model and simulation.
